---
title: "DEGs comparison in Day50 Cortical Brain Organoids"
author: "Oliviero Leonardi"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    Dataset: 'CHD2_iPSCs_and_organoids_PublicRepo'
    InputFolder: '/group/testa/Project/CHD2/BulkRNAseq/results/PublicRepo/2.DEA/'
    OutputFolder: '/group/testa/Project/CHD2/BulkRNAseq/results/PublicRepo/9.DEGsComparison/'
    FDRthr: 0.05
    logFCthr: 1.5

---
 
```{r EnvironmentSetupII, collapse=TRUE}
library(UpSetR)
library(RNASeqBulkExploratory)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(ggvenn)
library(gplots)
```

****

# 1. Environment Set Up

```{r EnvironmentSetupI, collapse=TRUE}
for (i in 1:length(params))
  print(paste('Parameter:', names(params)[i], ' - Value:', params[[i]], '- Class:', class(params[[i]])))
```

```{r collapse=TRUE}
logFCthr <- params$logFCthr
FDRthr <- params$FDRthr
```

## **DEGs are thresholded for Overlap Analysis:**

### __FDR threshold: `r params$FDRthr`__
### __LogFC threshold: `r params$logFCthr`__



# 2. Data Upload

## DEA for iPSCs

```{r, collapse=TRUE}
DEAList_ips_HT <- readRDS(paste0(params$InputFolder, 'iPSCs/Output/Savings/ipsc.DEAList_HT.rds'))
DEAList_ips_AR <- readRDS(paste0(params$InputFolder, 'iPSCs/Output/Savings/ipsc.DEAList_AR.rds'))

DEAList_ips_HT
DEAList_ips_AR
```

## DEA for Day25

```{r, collapse=TRUE}
DEAList_day25_HT <- readRDS(paste0(params$InputFolder, 'day25/Output/Savings/day25CbO.DEAList_HT.rds'))
DEAList_day25_AR <- readRDS(paste0(params$InputFolder, 'day25/Output/Savings/day25CbO.DEAList_AR.rds'))

DEAList_day25_HT
DEAList_day25_AR
```

## DEA for Day50


```{r, collapse=TRUE}
DEAList_day50_HT <- readRDS(paste0(params$InputFolder, 'day50/Output/Savings/day50CbO.DEAList_HT.rds'))
DEAList_day50_AR <- readRDS(paste0(params$InputFolder, 'day50/Output/Savings/day50CbO.DEAList_AR.rds'))

DEAList_day50_HT
DEAList_day50_AR
```

***

## Complete DEG List

### Already filtered DEGs from DEA

```{r, collapse=TRUE}
DEA <- list(D0_HT = row.names(DEAList_ips_HT$HT$DEGs),
            D0_AR = row.names(DEAList_ips_AR$AR$DEGs),
            D25_HT = row.names(DEAList_day25_HT$HT$DEGs), 
            D25_AR = row.names(DEAList_day25_AR$AR$DEGs),
            D50_HT = row.names(DEAList_day50_HT$HT$DEGs), 
            D50_AR = row.names(DEAList_day50_AR$AR$DEGs))

lapply(DEA, length)
```


### Manual filtering of DEGs

```{r, collapse=TRUE}
Res <- list(D0_HT = DEAList_ips_HT$HT$res, 
            D0_AR = DEAList_ips_AR$AR$res,
            D25_HT = DEAList_day25_HT$HT$res, 
            D25_AR = DEAList_day25_AR$AR$res,
            D50_HT = DEAList_day50_HT$HT$res,
            D50_AR = DEAList_day50_AR$AR$res)

Res <- lapply(Res, as.data.frame)

Res$D0_HT %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
Res$D0_AR %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
Res$D25_HT %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
Res$D25_AR %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
Res$D50_HT %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
Res$D50_AR %>% filter(., abs(log2FoldChange) > log2(logFCthr), padj < FDRthr)
```

****

# 3. Scatterplots

## iPSCs

```{r}
#Extract Results 
#NOTE: can extract more top genes using the Num argument, default Num=10
CompareResD0 <- computeCompResFC(Res$D0_HT, Res$D0_AR) 
```

```{r}
#Extract info on the genes belonging to the four quadrants
quadr <- c('Up_Up', 'Down_Down', 'Up_Down', 'Down_Up')

qDf <- lapply(quadr, {function(q) obj <- CompareResD0$Significant %>%
                                          dplyr::filter(Direction == q & Status == 'SignificantA&B') %>% 
                                          dplyr::select(genes, log2FoldChange_A, padj_A, 
                                                        log2FoldChange_B, padj_B, Log2FcSsum) %>%
                                          arrange(desc(abs(Log2FcSsum)))})
names(qDf) <- quadr
```

### Check DEGs overlap across time-point comparisons

```{r, collapse=TRUE}
compResFScatter(Res$D0_HT, Res$D0_AR)
```

Number of genes that are:

- Up_Up `r nrow(qDf[['Up_Up']])`
- Down_Down: `r nrow(qDf[['Down_Down']])`
- Up_Down: `r nrow(qDf[['Up_Down']])`
- Down_Up: `r nrow(qDf[['Down_Up']])`



### Concordant Downregulated genes

```{r}
qDf[['Down_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Discordant Upregulated and Downregulated genes

```{r}
qDf[['Up_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Discordant Downregulated and Upregulated genes

```{r}
qDf[['Down_Up']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

***

## Day25

```{r}
#Extract Results 
#NOTE: can extract more top genes using the Num argument, default Num=10
CompareResD25 <- computeCompResFC(Res$D25_HT, Res$D25_AR) 
```

```{r}
#Extract info on the genes belonging to the four quadrants
quadr <- c('Up_Up', 'Down_Down', 'Up_Down', 'Down_Up')

qDf <- lapply(quadr, {function(q) obj <- CompareResD25$Significant %>%
                                          dplyr::filter(Direction == q & Status == 'SignificantA&B') %>% 
                                          dplyr::select(genes, log2FoldChange_A, padj_A, 
                                                        log2FoldChange_B, padj_B, Log2FcSsum) %>%
                                          arrange(desc(abs(Log2FcSsum)))})
names(qDf) <- quadr
```

### Check DEGs overlap across time-point comparisons

```{r, collapse=TRUE}
compResFScatter(Res$D25_HT, Res$D25_AR)
```

Number of genes that are:

- Up_Up `r nrow(qDf[['Up_Up']])`
- Down_Down: `r nrow(qDf[['Down_Down']])`
- Up_Down: `r nrow(qDf[['Up_Down']])`
- Down_Up: `r nrow(qDf[['Down_Up']])`


### Concordant Upregulated genes

```{r}
qDf[['Up_Up']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Concordant Downregulated genes

```{r}
qDf[['Down_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Discordant Upregulated and Downregulated genes

```{r}
qDf[['Up_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

***

## Day50

```{r}
#Extract Results 
#NOTE: can extract more top genes using the Num argument, default Num=10
CompareResD50 <- computeCompResFC(Res$D50_HT, Res$D50_AR) 
```

```{r}
#Extract info on the genes belonging to the four quadrants
quadr <- c('Up_Up', 'Down_Down', 'Up_Down', 'Down_Up')

qDf <- lapply(quadr, {function(q) obj <- CompareResD50$Significant %>%
                                          dplyr::filter(Direction == q & Status == 'SignificantA&B') %>% 
                                          dplyr::select(genes, log2FoldChange_A, padj_A, 
                                                        log2FoldChange_B, padj_B, Log2FcSsum) %>%
                                          arrange(desc(abs(Log2FcSsum)))})
names(qDf) <- quadr
```

### Check DEGs overlap across time-point comparisons

```{r, collapse=TRUE}
compResFScatter(Res$D50_HT, Res$D50_AR)
```

Number of genes that are:

- Up_Up `r nrow(qDf[['Up_Up']])`
- Down_Down: `r nrow(qDf[['Down_Down']])`
- Up_Down: `r nrow(qDf[['Up_Down']])`
- Down_Up: `r nrow(qDf[['Down_Up']])`


### Concordant Upregulated genes

```{r}
qDf[['Up_Up']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Concordant Downregulated genes

```{r}
qDf[['Down_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Discordant Upregulated and Downregulated genes

```{r}
qDf[['Up_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Discordant Downregulated and Upregulated genes

```{r}
qDf[['Down_Up']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

***

## Day25 Heterozygous and Day50 Ancestral

```{r}
#Extract Results 
#NOTE: can extract more top genes using the Num argument, default Num=10
CompareResD25ht_50ar <- computeCompResFC(Res$D25_HT, Res$D50_AR) 
```

```{r}
#Extract info on the genes belonging to the four quadrants
quadr <- c('Up_Up', 'Down_Down', 'Up_Down', 'Down_Up')

qDf <- lapply(quadr, {function(q) obj <- CompareResD25ht_50ar$Significant %>%
                                          dplyr::filter(Direction == q & Status == 'SignificantA&B') %>% 
                                          dplyr::select(genes, log2FoldChange_A, padj_A, 
                                                        log2FoldChange_B, padj_B, Log2FcSsum) %>%
                                          arrange(desc(abs(Log2FcSsum)))})
names(qDf) <- quadr
```

### Check DEGs overlap across time-point comparisons

```{r, collapse=TRUE}
compResFScatter(Res$D25_HT, Res$D50_AR)
```

Number of genes that are:

- Up_Up `r nrow(qDf[['Up_Up']])`
- Down_Down: `r nrow(qDf[['Down_Down']])`
- Up_Down: `r nrow(qDf[['Up_Down']])`
- Down_Up: `r nrow(qDf[['Down_Up']])`


### Concordant Upregulated genes

```{r}
qDf[['Up_Up']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Concordant Downregulated genes

```{r}
qDf[['Down_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

### Concordant Downregulated genes

```{r}
qDf[['Up_Down']]%>% 
  DT::datatable(class='hover', rownames=FALSE, escape=TRUE, extension='Buttons', 
                options=list(dom='Bfrtip', buttons=list(I('colvis'), c('csv', 'excel'))))
```

***

# 4. upsetR {.tabset}

Check DEGs overlap across time-point comparisons.

## Heterozygous

```{r, collapse=TRUE, fig.width=8, fig.height=8, dpi=600}
upset(fromList(DEA[c(1,3,5)]), order.by = 'freq', nsets=6, matrix.color = 'darkgrey',
      keep.order = T, main.bar.color='grey10',
      decreasing = T, 
      mb.ratio = c(0.6, 0.4),
      text.scale = 1, 
      point.size = 3, 
      line.size = 1)
```

## Ancestral

```{r, collapse=TRUE, fig.width=8, fig.height=8, dpi=600}
upset(fromList(DEA[c(2,4,6)]), order.by = 'freq', nsets=6, matrix.color = 'darkgrey',
      keep.order = T, main.bar.color='grey10',
      decreasing = T, 
      mb.ratio = c(0.6, 0.4),
      text.scale = 1, 
      point.size = 3, 
      line.size = 1)
```

## All organoids

```{r, collapse=TRUE, fig.width=8, fig.height=8, dpi=600}
upset(fromList(DEA[c(3,4,5,6)]), order.by = 'freq', nsets=6, matrix.color = 'darkgrey',
      keep.order = T, main.bar.color='grey10',
      decreasing = T, 
      mb.ratio = c(0.6, 0.4),
      text.scale = 1, 
      point.size = 3, 
      line.size = 1)
```

## All samples

```{r, collapse=TRUE, fig.width=8, fig.height=8, dpi=600}
upset(fromList(DEA), order.by = 'freq', nsets=6, matrix.color = 'darkgrey',
      keep.order = T, main.bar.color='grey10',
      decreasing = T, 
      mb.ratio = c(0.6, 0.4),
      text.scale = 1, 
      point.size = 3, 
      line.size = 1)
```

# 5. Venn diagram

```{r fig.height=10, fig.width=8, dpi=300}
venn_plot <- ggvenn(DEA[3:6], show_elements=FALSE, fill_alpha = 0.2, text_size = 4, show_percentage = FALSE, fill_color = c('#FFDAB9', '#80CBC4', '#A0E8A9', '#C8A2C8'))

venn_plot

ggsave(paste0(params$OutputFolder, 'Output/Figures/', 'VennPlotOrganoids_1.pdf'))

venn_plot <- ggvenn(DEA[3:6], show_elements=FALSE, fill_alpha = 0.2, text_size = 4, show_percentage = FALSE, fill_color = c('#FFDAB9', '#80CBC4', '#FFDAB9', '#80CBC4' ))

venn_plot

ggsave(paste0(params$OutputFolder, 'Output/Figures/', 'VennPlotOrganoids_2.pdf'))

```

# 6. Intersection identities

```{r collapse=TRUE}
v.table <- venn(DEA[3:6])

all_intersections <- attr(v.table, "intersections")

for (comparison in names(all_intersections)) {
      cat(comparison, "\n")
      print(all_intersections[[comparison]])
}
```

*D25_AR:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_AR:D50_AR`
```

*D25_AR:D50_HT*
```{r collapse=TRUE}
all_intersections$`D25_AR:D50_HT`
```

*D25_HT:D25_AR*
```{r collapse=TRUE}
all_intersections$`D25_HT:D25_AR`
```

*D25_HT:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_HT:D50_AR`
```

*D25_HT:D50_HT*
```{r collapse=TRUE}
all_intersections$`D25_HT:D50_HT`
```

*D50_HT:D50_AR*
```{r collapse=TRUE}
all_intersections$`D50_HT:D50_AR`
```

*D25_AR:D50_HT:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_AR:D50_HT:D50_AR`
```

*D25_HT:D25_AR:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_HT:D25_AR:D50_AR`
```

*D25_HT:D25_AR:D50_HT*
```{r collapse=TRUE}
all_intersections$`D25_HT:D25_AR:D50_HT`
```

*D25_HT:D50_HT:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_HT:D50_HT:D50_AR`
```

*D25_HT:D25_AR:D50_HT:D50_AR*
```{r collapse=TRUE}
all_intersections$`D25_HT:D25_AR:D50_HT:D50_AR`
```


## Gene Ontology on genes shared by Day25 HT and Day50 AR

```{r fig.width=8, fig.height=4}
D25_HT_D50_AR_overlapGenes <- all_intersections$`D25_HT:D50_AR` #Genes in overlap

Universe <- union(rownames(DEAList_day25_HT$HT$res), rownames(DEAList_day50_AR$AR$res)) #Union of detected genes for universe

GeneVectors <- list()
GeneVectors$DEGenes <- ifelse(Universe %in% D25_HT_D50_AR_overlapGenes, 1, 0)
names(GeneVectors$DEGenes) <- Universe


BPann <- topGO::annFUN.org(whichOnto="BP", feasibleGenes=names(GeneVectors$DEGenes), 
                    mapping="org.Hs.eg.db", ID="symbol") %>% inverseList()

ResBP_D25_HT_D50_AR_overlapGenes <- topGOResults(Genes=GeneVectors$DEGenes, gene2GO=BPann, ontology='BP', geneSelectionFun = function(x) (x == 1),
                         desc=NULL, nodeSize=3, algorithm='weight01', statistic='fisher', 
                         EnTh=1, PvalTh=0.05, minTerms=12, geneTh=3,
                         saveRes=FALSE)
```


```{r}
custom_theme <- theme(
  panel.background = element_blank(),  # No background color
  panel.border = element_rect(color = "black", fill = NA, size = 0.5),  # Black border around panels
  panel.grid.major = element_line(color = "grey90", size = 0.5),  # Grey major grid lines
  panel.grid.minor = element_blank(),  # No minor grid lines
  axis.text = element_text(color = "black"),  # Black text for axis labels
  axis.title = element_text(color = "black"),  # Black text for axis titles
  plot.title = element_text(color = "black", hjust = 0.5),  # Black plot title, centered
  legend.position = "top",  # No legend
  strip.background = element_blank(),  # No background for facet labels
  strip.text = element_text(color = "black", face = "bold")  # Black, bold text for facet labels
)
```

```{r fig.width=8, fig.height=4}
ResOrdered <- ResBP_D25_HT_D50_AR_overlapGenes$ResSel
ResOrdered <- ResOrdered[order(ResOrdered$Statistics, decreasing = TRUE), ]
MaxVal <- round(max(-log10(as.numeric(ResOrdered$Statistics)), na.rm=TRUE), 0) +1
ResOrdered$fillcolor <- '#ed72a9'# '#31688EFF'
ResOrdered$Term <- factor(ResOrdered$Term, levels = ResOrdered$Term)

ResOrdered <- ResOrdered %>% filter(ResOrdered$Statistics < 0.05)

ggplot(data=ResOrdered, aes(x=Term, y=-log10(as.numeric(Statistics)), size=ER)) + 
    geom_hline(yintercept=-log10(0.05), col='gray80', lty='longdash') +
    geom_point(stat='identity', colour="black", pch=21, fill=ResOrdered$fillcolor) +
    #geom_text(aes(y=0), label=ResOrdered$Term, hjust=0, size=4, color='black') + 
    scale_y_continuous(breaks=seq(0,MaxVal,2), labels=abs(seq(0, MaxVal, 2)), limits=c(0,MaxVal), expand=c(0.025, 0.025)) +
    ylab('-log10 PValue') + xlab('') +
    theme_bw() +
    labs(size='Enrichment Ratio') +
    theme(legend.position='right', axis.text = element_text(colour = 'black', size=12), 
          plot.title= element_text(face='', size=10, hjust = 0.5)) +
    custom_theme +
    scale_x_discrete(position = "top") +
    coord_flip()


ggsave(paste0(params$OutputFolder, 'Output/Figures/', 'GO_BP_sharedGenes_D25HTandD50AR.pdf'), device = 'pdf', width = 8, height = 4)
```

## Gene Ontology on genes in Day25 HT and Day50 Overlap with SFARI genes
```{r}
D25_HT_D50_AR_overlapGenesSFARI <- rownames(CompareResFlip$Significant)
GeneVectors$DEGenesSFARI <- ifelse(Universe %in% D25_HT_D50_AR_overlapGenesSFARI, 1, 0)
names(GeneVectors$DEGenesSFARI) <- Universe


ResBP_D25_HT_D50_AR_overlapGenesSFARI <- topGOResults(Genes=GeneVectors$DEGenesSFARI, gene2GO=BPann, ontology='BP', geneSelectionFun = function(x) (x == 1),
                         desc=NULL, nodeSize=1, algorithm='weight01', statistic='fisher', 
                         EnTh=1, PvalTh=0.05, minTerms=12, geneTh=1,
                         saveRes=FALSE)

## No GO category enriched
```

***

```{r collapse=TRUE}
sessionInfo()
``` 


***

### **`r paste0('last update on: ', date())`**

***





