---
title: "AutophagyComplexHeatmap All"
author: "Oliviero Leonardi"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document

params: 
    AutophagyTable: '/group/testa/Project/CHD2/BulkRNAseq/data/GeneSets/Autophagy.txt'
    ht.day25.deseq: '/group/testa/Project/CHD2/BulkRNAseq/results/PublicRepo/2.DEA/day25/Output/Savings/day25CbO.deseqHTvsWT.rds'
    ht.day50.deseq: '/group/testa/Project/CHD2/BulkRNAseq/results/PublicRepo/2.DEA/day50/Output/Savings/day50CbO.deseqHTvsWT.rds'
    FiguresFolder: '/group/testa/Project/CHD2/BulkRNAseq/results/PublicRepo/3.Signatures/Output/Figures/'
    padj_threshold: 0.05
    logFC_threshold: no
    logFC_ceiling: 2
---

```{r EnvironmentSetupI, include=FALSE}
for (i in 1:length(params))
  print(paste('Parameter:', names(params)[i], ' - Value:', params[[i]], '- Class:', class(params[[i]])))
```

```{r include=FALSE}
#Define FDR threshold
padj_threshold <- params$padj_threshold

#Define logFC threshold
logFC_thrshold <- params$logFC_threshold

#Define ceiling value accoriding to params
ceiling_value <- params$logFC_ceiling
```

***

## Thresholds

**FDR threshold is set to:** `r padj_threshold`

**LogFC threshold is set to:** NULL

**LogFC ceiling is set to:** `r ceiling_value`

***

### Libraries

```{r include=FALSE}
library(ComplexHeatmap)
library(dplyr)
library(SEtools)
library(circlize)
library(ggplot2)
```


### Import DEGs from DESeq2

```{r, collapse=TRUE}
degs_ht_day25 <- data.frame(readRDS(params$ht.day25.deseq))[, c(2,6)]
degs_ht_day50 <- data.frame(readRDS(params$ht.day50.deseq))[, c(2,6)]

degs_ht_day25 <- tibble::rownames_to_column(degs_ht_day25, var = 'GeneSymbol')
degs_ht_day50 <- tibble::rownames_to_column(degs_ht_day50, var = 'GeneSymbol')

degs_ht_day25 <- degs_ht_day25 %>% filter(padj < padj_threshold)
degs_ht_day50 <- degs_ht_day50 %>% filter(padj < padj_threshold)

degs_ht_day25$dataset <- 'Day25_Heterozygous_Organoids'
degs_ht_day50$dataset <- 'Day50_Heterozygous_Organoids'
```

```{r}
degs <- purrr::reduce(list(degs_ht_day25, degs_ht_day50), full_join, by = "GeneSymbol")

degs <- degs %>%
  rename(log2FoldChange_day25CbO_Ht = log2FoldChange.x, padj.day25CbO_Ht = padj.x,
         log2FoldChange_day50CbO_Ht = log2FoldChange.y, padj.day50CbO_Ht = padj.y)
```

### Autophagy signature genes

```{r, collapse=TRUE}
#Import Autophagy table
AutophagyDF <- read.table(params$AutophagyTable, sep = '\t', header=TRUE)
```

### Annotation dataframe

```{r, collapse=TRUE}
#Make annotation df from autophagy dataframe
annotation_df <- data.frame(GeneSymbol=AutophagyDF$GeneSymbol,
                            Category=AutophagyDF$Category)

#Remove duplicates
#annotation_df <- annotation_df[!duplicated(annotation_df$GeneSymbol), ]
```

### Merge DEGs and Autophagy signature df

```{r, collapse=TRUE}
#Make degs dataframe with intersection of DEGs and AutophagyGenes
degs <- left_join(degs, annotation_df, by="GeneSymbol")

#Remove NAs
degs <- degs[!is.na(degs$Category), ]
```

### Generate matrix

```{r, collapse=TRUE}
#Generate matrix for heatmap with row and column names
mat <- as.matrix(degs %>% dplyr::select(starts_with("log2FoldChange")))
rownames(mat) <- degs$GeneSymbol

head(mat)
```

```{r}
columNamesVector <- c('Day25CbO_HT','Day50CbO_HT')
```


### Split categories 

```{r, collapse=TRUE}
#Lysosome
lysogenes <- degs[degs$Category == 'Lysosome', ]
lysomat <- as.matrix(lysogenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(lysomat) <- lysogenes$GeneSymbol
colnames(lysomat) <- columNamesVector
lysomat[is.na(lysomat)] <- 0


#Core autophagy genes
coregenes <- degs[degs$Category == 'Autophagy core', ]
coremat <- as.matrix(coregenes%>% dplyr::select(starts_with("log2FoldChange")))
rownames(coremat) <- coregenes$GeneSymbol
colnames(coremat) <- columNamesVector
coremat[is.na(coremat)] <- 0


#mTOR and upstream pathways
mtorgenes <- degs[degs$Category == 'mTOR and upstream pathways', ]
mtormat <- as.matrix(mtorgenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(mtormat) <- mtorgenes$GeneSymbol
colnames(mtormat) <- columNamesVector
mtormat[is.na(mtormat)] <- 0


#Autophagy regulators
regolatorgenes <- degs[degs$Category == 'Autophagy regulators', ]
regolatormat <- as.matrix(regolatorgenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(regolatormat) <- regolatorgenes$GeneSymbol
colnames(regolatormat) <- columNamesVector
regolatormat[is.na(regolatormat)] <- 0


#Mitophagy
mitophagygenes <- degs[degs$Category == 'Mitophagy', ]
mitophagymat <- as.matrix(mitophagygenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(mitophagymat) <- mitophagygenes$GeneSymbol
colnames(mitophagymat) <- columNamesVector
mitophagymat[is.na(mitophagymat)] <- 0


#Docking and fusion
dockingenes <- degs[degs$Category == 'Mitophagy', ]
dockingmat <- as.matrix(dockingenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(dockingmat) <- dockingenes$GeneSymbol
colnames(dockingmat) <- columNamesVector
dockingmat[is.na(dockingmat)] <- 0


#Lysosome-related
lysorelatedgenes <- degs[degs$Category == 'Mitophagy', ]
lysorelatedmat <- as.matrix(lysorelatedgenes %>% dplyr::select(starts_with("log2FoldChange")))
rownames(lysorelatedmat) <- lysorelatedgenes$GeneSymbol
colnames(lysorelatedmat) <- columNamesVector
lysorelatedmat[is.na(lysorelatedmat)] <- 0
```

```{r, collapse=TRUE}
#Category apping color
annotation_colors <- c("mTOR and upstream pathways" = "red",
                     "Autophagy core" = "blue",
                     "Autophagy regulators" = "green",
                     "Mitophagy" = "orange",
                     "Docking and fusion" = "purple",
                     "Lysosome" = "yellow",
                     "Lysosome-related" = "pink")

annotation_colors <- c("mTOR and upstream pathways" = "#A2F07B",
                     "Autophagy core" = "#EB489C",
                     "Autophagy regulators" = "#8D6AFF",
                     "Mitophagy" = "#EB5E21",
                     "Docking and fusion" = "#F0DA2A",
                     "Lysosome" = "#2C7FB8",
                     "Lysosome-related" = "#70E6CC")
```

### Color function
```{r, collapse=TRUE}
#Color function
col_fun = colorRamp2(c(-2, 0, 2), c("#3b528b", "white", "#f1605d"))
col_fun(seq(-3, 3))
```

### Set rowannotation color

```{r, collapse=TRUE}
#RowAnnotation
rowcolor <- annotation_colors[names(annotation_colors) %in% degs$Category]
category_anno <- rowAnnotation(Category = degs$Category, 
                               col = list(Category = rowcolor))
```


### Set LogFC ceiling to matrix

```{r fig.height=10, fig.width=5}
#Set upper and lower threshold values
threshold_upper <- ceiling_value
threshold_lower <- -ceiling_value

#Replace values exceeding the thresholds
mat[mat > threshold_upper] <- threshold_upper
mat[mat < threshold_lower] <- threshold_lower
```


### Heatmap with all categories

```{r fig.height=40, fig.width=5}
allmat <- as.matrix(degs %>% dplyr::select(starts_with("log2FoldChange")))
rownames(allmat) <- degs$GeneSymbol
colnames(allmat) <- columNamesVector
allmat[is.na(allmat)] <- 0

Heatmap(allmat, col = col_fun, name = 'log2FoldChange', right_annotation = category_anno,
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(1, "cm"))
```

### Heatmap for lysosomes

```{r fig.height=9, fig.width=10}
Heatmap(lysomat, col = col_fun, name = 'log2FoldChange',
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(1, "cm"))

pdf(paste0(params$FiguresFolder, 'Lyso_heatmap_het_organoids.pdf'), width = 7, height = 10)
Heatmap(lysomat, col = col_fun, name = 'log2FoldChange', border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(0.75, "cm"), row_names_rot = 0, 
        row_names_gp = gpar(fontsize = 11, rot = 0))
dev.off()
```

### Heatmap for core autophagy genes

```{r fig.height=13, fig.width=10}
Heatmap(coremat, col = col_fun, name = 'log2FoldChange',
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(1, "cm"))
```

### Heatmap for autophagy regulators

```{r fig.height=9, fig.width=10}
Heatmap(regolatormat, col = col_fun, name = 'log2FoldChange',
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(1, "cm"))
```

### Heatmap with mTOR and usptream pathways
```{r fig.height=9, fig.width=10}
Heatmap(mtormat, col = col_fun, name = 'log2FoldChange',
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(1, "cm"))
```

### Fraction of genes in categories

```{r, collapse=TRUE}
#Compute fraction of genes in each category
fraction <- data.frame(table(degs$Category) / length(degs$Category))

ggplot(fraction, aes(x=Freq, y=Var1)) +
  geom_col() + theme_bw()
```

### Heatmap for lysosomes and autophagy core genes
```{r fig.height=23, fig.width=7}
print('genes in common:')
intersect(rownames(lysomat), rownames(coremat))

#drop TMEM59 from lysomat; then the annotation will be fixed in illustrator making a single square with both color

toexclude <- intersect(rownames(lysomat), rownames(coremat))
lysomat <- lysomat[!rownames(lysomat) %in% toexclude, , drop = FALSE]

lysocoremat <- rbind(lysomat, coremat)

x <- c('Lysosomes', 'Autophagy core genes')

rows <-  rep(x, times=c(nrow(lysomat), nrow(coremat)))
row_annotation <- HeatmapAnnotation(gp = gpar(col = "black"),
                                    Category = rows, which = "row", width = unit(1, "cm"),
                                    col=list(Category = c('Lysosomes' = '#22a884', 
                                                          'Autophagy core genes' = '#453781')))

lysocoreheatmap <- Heatmap(lysocoremat, col = col_fun, name = 'log2FoldChange', left_annotation = row_annotation,
        border = 'black', rect_gp = gpar(col = "black", lwd = 1), 
        width = unit(0.75, "cm"), row_names_rot = 0, 
        row_names_gp = gpar(fontsize = 11, rot = 0))

lysocoreheatmap

pdf(paste0(params$FiguresFolder, 'LysoAndAutophagyCoreGenes_heatmap.pdf'), width = 7, height = 23)
lysocoreheatmap
dev.off()
```


***

### **`r paste0('last update on: ', date())`**

***
